### 2021-04-20 introduction.tex
Update introduction.tex

In the $\pi$-calculus literature, there is a growing line of work in developing Curry-Howard correspondences between session-typed $\pi$-calculus and linear logic~\cite{girard87}: Caires and Pfenning's $\pi$DILL~\cite{cairespfenning10} corresponds to dual intuitionistic linear logic~\cite{barber96}, and Wadler's Classical Processes~\cite[CP]{wadler14} corresponds to classical linear logic~\cite[CLL]{girard87}. Both calculi guarantee deadlock freedom, which they achieve by restricting structure of processes and shared channels to \emph{trees}, by combing name restriction and parallel composition into a single construct, corresponding to the logical cut. This ensures that two processes can only communicate via exactly one series of channels, which rules out interleavings of sessions, and guarantees deadlock freedom.
There are many downsides to combining name restriction and parallel composition---lack of modularity, difficulty typing structural congruence and formulating label-transition semantics, \etc---which have led to various approaches to decoupling these constructs. Hypersequent CP~\cite{MP18,kokkemontesi19popl,kokkemontesi19tlla} and Linear Compositional Choreographies~\cite{CarboneMS18} decouple them, but maintain the correspondence to CLL and allow only tree-structured processes. Priority CP~\cite[PCP]{dardhagay18extended} weakens the correspondence to CLL in exchange for a more expressive language which allows cyclic-structured processes. PCP decouples CP's cut rule into two separate constructs: one for parallel composition via a mix rule, and one for name restriction via a cycle rule. To restore deadlock freedom, PCP uses \emph{priorities}~\cite{kobayashi06,padovani14}. Priorities encode the \emph{order of actions} and rule out bad interleavings. Dardha and Gay~\cite{dardhagay18extended} prove cycle-elimination for PCP, adapting the cut-elimination proof for classical linear logic, and deadlock freedom follows as a corollary.

CP and GV are related via a pair of translations which satisfy simulation~\cite{lindleymorris16}, and which can be tweaked to satisfy reflection. The two calculi share the same strong guarantees. GV achieves deadlock freedom via a similar syntactic restriction: it combines channel creation and thread spawning into a single operation, called ``fork'', which is related to the cut construct in CP. Unfortunately, as with CP, this syntactic restriction has many downsides.

Our aims are to develop the expressiveness of GV but maintain deadlock freedom. We chose GV for several reasons, some of which apply more generally as benefits to working within a concurrent $\lambda$-calculus as opposed to a name-passing process calculus. Concurrent $\lambda$-calculi support higher-order functions, and have a capability for abstraction not usually present in process calculi. Within a concurrent $\lambda$-calculus, one can derive extensions of the communication capabilities of the language via well-understood extensions of the functional fragment, \eg, we can derive internal/external choice from sum types. Concurrent $\lambda$-calculi maintain a clear separation between the program which the user writes and the configurations which represent the state of the system as it evaluates the program. However, our main motivation is that results obtained for $\lambda$-calculi transfer more easily to real-world functional programming languages. \emph{Case in point}: we were able to embed the theory developed in this paper in Linear Haskell~\cite{bernardyboespflug18} without much effort~\cite{KD21b}. For us, these advantages outweigh the benefits of process calculi, \eg, their succinctness.
The benefits of working with GV specifically, as opposed to other concurrent $\lambda$-calculi, is its relation to CP~\cite{wadler14}, and its strong formal properties. We thus pose our research question for GV:
\begin{quotation}
  \textbf{RQ:}
  Can we design a more expressive GV which guarantees deadlock freedom for cyclic-structured processes?
\end{quotation}

>>>>

In the $\pi$-calculus literature, there is a growing line of work in developing Curry-Howard correspondences between session-typed $\pi$-calculus and linear logic~\cite{girard87}: Caires and Pfenning's $\pi$DILL~\cite{cairespfenning10} corresponds to dual intuitionistic linear logic~\cite{barber96}, and Wadler's Classical Processes~\cite[CP]{wadler14} corresponds to classical linear logic~\cite[CLL]{girard87}. Both calculi guarantee deadlock freedom, which they achieve by restricting structure of processes and shared channels to \emph{trees}, by combing name restriction and parallel composition into a single construct, corresponding to the logical cut. This ensures that two processes can only communicate via exactly one series of channels, which rules out interleavings of sessions, and guarantees deadlock freedom.
There are many downsides to combining name restriction and parallel composition, such as lack of modularity, difficulty typing structural congruence and formulating label-transition semantics, which have led to various approaches to decoupling these constructs. Hypersequent CP~\cite{MP18,kokkemontesi19popl,kokkemontesi19tlla} and Linear Compositional Choreographies~\cite{CarboneMS18} decouple them, but maintain the correspondence to CLL and allow only tree-structured processes. Priority CP~\cite[PCP]{dardhagay18extended} weakens the correspondence to CLL in exchange for a more expressive language which allows cyclic-structured processes. PCP decouples CP's cut rule into two separate constructs: one for parallel composition via a mix rule, and one for name restriction via a cycle rule. To restore deadlock freedom, PCP uses \emph{priorities}~\cite{kobayashi06,padovani14}. Priorities encode the \emph{order of actions} and rule out bad cyclic interleavings. Dardha and Gay~\cite{dardhagay18extended} prove cycle-elimination for PCP, adapting the cut-elimination proof for classical linear logic, and deadlock freedom follows as a corollary.

CP and GV are related via a pair of translations which satisfy simulation~\cite{lindleymorris16}, and which can be tweaked to satisfy reflection. The two calculi share the same strong guarantees. GV achieves deadlock freedom via a similar syntactic restriction: it combines channel creation and thread spawning into a single operation, called ``fork'', which is related to the cut construct in CP. Unfortunately, as with CP, this syntactic restriction has its downsides.

Our aim is to develop a more expressive version of GV while maintaining deadlock freedom. We chose GV for several reasons, some of which apply more generally as benefits to working within a concurrent $\lambda$-calculus as opposed to a name-passing process calculus. Concurrent $\lambda$-calculi support higher-order functions, and have a capability for abstraction not usually present in process calculi. Within a concurrent $\lambda$-calculus, one can derive extensions of the communication capabilities of the language via well-understood extensions of the functional fragment, \eg, we can derive internal/external choice from sum types. Concurrent $\lambda$-calculi maintain a clear separation between the program which the user writes and the configurations which represent the state of the system as it evaluates the program. However, our main motivation is that results obtained for $\lambda$-calculi transfer more easily to real-world functional programming languages. Case in point: we embed the theory developed in this paper in Linear Haskell~\cite{bernardyboespflug18} where PGV easily translates into a library for deadlock free session typed programming~\cite{kokkedardha21hs}. %For us, these advantages outweigh the benefits of process calculi, \eg, their succinctness.
The benefit of working specifically with GV, as opposed to other concurrent $\lambda$-calculi, is its relation to CP~\cite{wadler14}, and its formal properties, including deadlock freedom. We thus pose our research question for GV:
\begin{quotation}
  \textbf{RQ:}
  Can we design a more expressive GV which guarantees deadlock freedom for cyclic-structured processes?
\end{quotation}
